Reproducibility and Open Science in Economics

### 2025-03-10 article.tex
Updates from Overleaf

@@

A key ambiguity will arise in how big such networks need to be in order to be considered ``open.'' Clearly, $n=2$ is not considered a network. The \ac{NBER} defines its affiliated scholars as a network: $n=1804$ as of January 2025, primarily in North America \citep{national_bureau_of_economic_research_affiliated_2025}. However, in \nberyear{}, a total of \nberauthorsunique{} authors published \nberpapers{} NBER working papers. J-PAL has approximately $n=1725$, at 120 universities on all populated continents \citep{abdul_latif_jameel_poverty_action_lab_affiliated_2025}. Between 2001 and June 2024, there had been $n=2023$ researchers on projects that used confidential U.S. Census Bureau in the \ac{FSRDC} \citep{us_census_bureau_uscensusbureaufsrdc-external-census-projects_2024}. In an average year (2013-2023), $n=\economicsgrads{}$ students graduate from a U.S. university with a Ph.D. in economics \citep[Table 1-5]{national_science_foundation_doctorate_2024}. In the approximately 30 years since inception of \ac{RePEc}, $n=526$ authors from 52 institutions in \textbf{Norway} have published a paper listed on \ac{RePEc} (presumably in economics) \citep{ideasrepec_within_2025}. All of these are measured across different spatio-temporal dimensions. Are they large? Context and purpose matter. Some may intersect. How many U.S. graduate students in the past 10 years have published an NBER working paper and are now at a Norwegian institution?

@@

Many well-known survey organizations impose redistribution and usage restrictions that are not consistent with the OKD definition. Many such redistribution restrictions apply to datasets with more detailed personal information collected through surveys, and are meant to ensure continued compliance with ethical rules of behavior, often with informed consent agreements by survey participants and local privacy laws. Notable examples include  PSID \parencite{institute_for_social_research_panel_2024}, World Value Survey \parencite{haerpfer_world_2024}, Demographic Household Surveys \parencite{dhs_program_demographic_2024}, German Socio-Economic Panel \parencite{goebel_german_2019,goebel_socio-economic_2024}, all of which have broad (cost-free) usage, conditional on registration and compliance with usage restrictions. Table~\ref{tab:restrictions} shows a few examples.

@@

The last three rows of  Table~\ref{tab:restrictions}, however, go further. Researchers using SOEP data must satisfy certain geographical requirements, such as presence of the researcher in the countries covered by \ac{GDPR}, otherwise they can only access a modified version of the data. Similar restrictions apply to the restricted-use data of many other surveys, for instance, the US-based \ac{NLSY}. Users of restricted PSID data must comply with even more stringent rules: secure approval from their institution's ethics board, and use secure computing environments. Clearly, these restrictions will inhibit broader use of the data per se. Yet these restrictions are not globally very restrictive: There are in Economics alone \economicsgrads{} new US-based researchers being given the ability to request access to geo-restricted NLSY data every year. A similar number in the EU likely obtain the right to access the geo-restricted SOEP data as well. Despite the restrictions on the use of SOEP data, there are currently \soeppapers{} that in some fashion use the data.\footnote{Source: \url{https://www.diw.de/en/diw_01.c.789503.en/publications_based_on_soep_data__soeplit.html} accessed on 2025-02-08.} 
The PSID bibliography\footnote{Source: \url{https://psidonline.isr.umich.edu/publications/Bibliography/search.aspx} accessed on 2025-02-08} lists over 1,300 dissertations and over 5,300 articles.
In the space depicted in Figure~\ref{fig:nxt}, access is very much towards the left of the figure.%
%
\footnote{In fact, authors sometimes forget to abide by the rules for these datasets. As AEA Data Editor, I get notified via ``take-down requests'' from data providers 12-15 times per year, including in 2024 from the PSID, and have posted information on how to achieve compliance, at least in some cases, at \url{https://aeadataeditor.github.io/posts/2024-11-01-psid-requests}. Most of the cases affect papers published prior to my tenure, because I do alert authors to  data use agreement violations that I am aware of. Ultimately, however, it is the authors' obligation to remain compliant with such data use agreements.}.

\paragraph{Organic data:} However, most economists, when asked about data subject to restrictions, will think of ``proprietary data,'' a term often applied to any data that may be subject to restrictions of use and re-use. Access to most administrative data is typically not ``open'' in the sense of OKD, but are they open enough, given the privacy concerns that are attached to these data? Many have argued that access is not broad enough \citep{card_expanding_2010,einav_economics_2014}, while acknowledging the difficulty of addressing privacy and security of the data at scale. \citet{abowd_economic_2018} discuss the challenge of making the choice of between accuracy of (public) statistics and data, and the privacy loss inherent in doing so. \citet{nagaraj_improving_2020} argue that more openness improves scientific progress, and \citet{nagaraj_how_2023} study this in the context of the US system for providing access (\ac{FSRDC}). They note that 4\% of US-based empirical authors have had some access to the FSRDC system. I am not aware of similar studies for other countries, such as France \citep[the equivalent system is the \ac{CASD}, ][]{gadouche_centre_2019} or Canada \citep{currie2015social}. In absolute terms, these networks host several hundred researchers every year. As per (CENSUS), 2,084 researchers have had access to projects involving Census Bureau data since (YEAR).\footnote{\url{https://labordynamicsinstitute.github.io/fsrdc-external-census-projects/}}. \citet{nagaraj_how_2023} mention 861 papers. Similar numbers can be obtained for the French (xxx researchers and 417 publications)
% https://www.casd.eu/toutes-les-publications/liste/0/20/ 
% 20 pages à 20 publications et 1 à 17 = 417
and Canadian (2201  active researchers as of January 2025, and 3245 papers published between 2000 and May 2024)\footnote{Provided by Grant Gibson, CRDCN, on February 10, 2025.} networks.\footnote{The \citet{nagaraj_how_2023} number only includes papers published by economists. All other numbers are counts of researchers and publications in all disciplines, and include non-economists.} The number of publications from access to these networks is smaller in absolute terms then those from PSID and SOEP, though likely higher in impact \citep{nagaraj_how_2023}.

\paragraph{Primary data collection:} The discussion of choices made by survey organizations should in principle be applicable when smaller teams of economists, not entire survey organizations, do the primary data collection. Similar to survey organizations, such teams have to balance the privacy of their respondents with the benefits of open science, in particular the broader knowledge to be gained from open access to the data. Many, so it would seem, provide much of the data in replication packages, subject  to de-identification \citep[see ][ for examples]{kopper_j-pal_2020,bjarkefur_development_2021}, though typically not with stronger disclosure avoidance measures similar to those employed by statistical agencies and larger survey institutions \citep[for a brief discussion of the issues and one possible solution, see ][]{mukherjee_assessing_2023}. For research teams, ethics boards and \acp{IRB} have a role to play \citep{grant_opinion_2019}, with some arguing very strongly that greater availability to others (though not blind publication of all data) is required in order to maximize the societal benefits that are the \textit{quid pro quo} for the respondents' consent to their privacy being invaded \citep{meyer_practical_2018,grant_opinion_2019}. Making such data as broadly available, while respecting the privacy of respondents, is precisely what open access to such data promises, \textit{modulo} appropriate access restrictions or data use agreements similar to those outlined in Table~\ref{tab:restrictions}. In general, however, primary data collections do not have access to robust third-party systems that would allow for access similar to the access required by PSID and similar organizations, situated between no access and fully public access. Thus, while access may be requested in ad-hoc fashion via the original authors, this is known to be fraught with problems \citep{watson_many_2022,gabelica_many_2022}. An ideal scenario would see researchers deposit the data they collected in third-party repositories, which then handle issues such as verifying ethics approval and secure access mechanisms. Some full-service repositories, such as \ac{ICPSR} or various national archives, offer such deposits, though they are rarely used by individual researchers. 
%
% Swedish example
% Das paper ist
One example are the data collected and used by \citet{ahrsjo_identity_2024}. The data were collected from public information from the Stockholm District Court. However, in combining individuals' information into a database, the result was subject to \ac{GDPR}, and could not be included in the replication package \citep{ahrsjo_code_2024}. 
%Ahrsjö, Ulrika, Susan Niknami, and Mårten Palme. “Identity in Court Decision-Making.” American Economic Journal: Economic Policy 16, no. 4 (November 2024): 142–64. https://doi.org/10.1257/pol.20220802.
%
% Ahrsjö, Ulrika, Niknami, Susan, and Palme, Mårten. Code for Identity in Court Decision-Making. Nashville, TN: American Economic Association [publisher], 2024. Ann Arbor, MI: Inter-university Consortium for Political and Social Research [distributor], 2024-10-07. https://doi.org/10.3886/E193364V1
% Im Kontext, die Autoren bauen einen "data set created from court documents from the Stockholm District Court." Den dürfen sie aber nicht, laut ihrer Uni, wegen GDPR veröffentlichen. Also haben sie ihn beim "Swedish National Data Service" hinterlegt:
%
% Ahrsjö, U., Niknami, S., & Palme, M. (2023). Data för: Identity in Court Decision-Making (Version 1) [Dataset]. Stockholms universitet. Tillgänglig via: https://doi.org/10.58141/27j7-8606
However, the authors were able to deposit their pseudoymous data at the Svensk nationell datatjänst (Swedish National Data Service), as a restricted dataset \citep{ahrsjo_data_2023}. The SNDS will verify compliance with Swedish law (e.g., \ac{GDPR}), without requiring future involvement of the original collectors of the data.

How onerous are access restrictions in general? In my work as Data Editor, it matters less than it would seem at first blush. The AEA asks that authors, when submitting, provide some information on how restrictive the data used in the article are, and whether the authors are able to provide the data editor with a not-for-publication copy, for the purpose of verification.\footnote{\href{https://www.aeaweb.org/journals/forms/data-code-availability}{Data and Code Availability Form}} Table~\ref{tab:categories} lists the four levels of restrictions, with some guidance on how to classify specific data. Access via research data centers, described earlier, generally fit within the `moderately difficult to obtain` category. Many private-sector datasets, relying on personal interactions with individuals within the company, fall into the `very difficult to obtain', because others may never be able to obtain them. Authors can list multiple categories, and we review these, possibly adjusting them before we record them in our internal database.

@@

Table~\ref{tab:access_provision} lists the distribution of the four levels of restrictions for the 96 AER articles in 2024, in addition to the no-restriction category, interacted with whether the authors offered to share the files with the data editor.\footnote{Not all offers to share data are taken up, depending on available resources and time.}  Slightly more than 50\% of all articles had no data restrictions, which is a bit lower than long-term averages. However, the combination of no access restriction and `very easy to obtain' (about 68 out of 96, 70\%), is higher than the usual quoted number of around 60\% of access-free data \parencite{herbert_reproduce_2024,einav_economics_2014}. More importantly, out of the 39 data sources that are difficult to obtain, about a third could actually be shared with the data editor. It is also worthwhile highlighting that the `moderately difficult to obtain' tends to be data from research data center networks that have fairly robust mechanisms of access, so that in principle and in practice, many researchers do have access to the same data used in these articles. Arguably, the only data sources that are of immediate concern are data where access is uniquely attributable to the authors of the paper, or where access is no longer possible (e.g., Zillow's ZTRAX program, terminated in 2021).

Concern should be raised, however, for any data that is subject to a gatekeeping mechanism, however objective, impartial, and lightweight it may be at this time, as such mechanisms can and do disappear.  A particular striking example, as of the writing of this article, is access to data from the USAID-funded DHSprogram. Data access of such data was classified as `moderately easy to obtain' in Table~\ref{tab:categories} until February 2025, as it took only a day or two to obtain access subject to a lightweight data use agreement. In February 2025, the second Trump administration shut down USAID and the DHS Program, with very little notice. While the system remained (for a time) accessible to those with prior access permission, no further access requests were accepted. That turns it into a `very difficult to obtain' dataset.\footnote{As of March 2025, efforts are underway to preserve the DHS program data, for instance via IPUMS, and to resuscitate both the access to historical data, as well continue funding new surveys.} These events are a note of caution that any kind of redistribution restriction may very well negatively impact future availability to the research community.

@@

\printbibliography[title={References}]
\end{document}

>>>>

% bibliography files

\addbibresource{from-zotero.bib}
\addbibresource{references.bib} % generated dynamically by Overleaf

@@

A key ambiguity will arise in how big such networks need to be in order to be considered ``open.'' Clearly, $n=2$ is not considered a network. The \ac{NBER} defines its affiliated scholars as a network: $n=1804$ as of January 2025, primarily in North America \citep{national_bureau_of_economic_research_affiliated_2025}. However, in \nberyear{}, a total of \nberauthorsunique{} authors published \nberpapers{} NBER working papers. J-PAL has approximately $n=1725$ affiliates at 120 universities on all populated continents \citep{abdul_latif_jameel_poverty_action_lab_affiliated_2025}. Between 2001 and June 2024, there had been $n=2023$ researchers on projects that used confidential U.S. Census Bureau in the \ac{FSRDC} \citep{us_census_bureau_uscensusbureaufsrdc-external-census-projects_2024}. In an average year (2013-2023), $n=\economicsgrads{}$ students graduate from a U.S. university with a Ph.D. in economics \citep[Table 1-5]{national_science_foundation_doctorate_2024}. In the approximately 30 years since inception of \ac{RePEc}, $n=526$ authors from 52 institutions in \textbf{Norway} have published a paper listed on \ac{RePEc} (presumably in economics) \citep{ideasrepec_within_2025}. All of these are measured across different spatio-temporal dimensions. Are they large? Context and purpose matter. Some may intersect. How many U.S. graduate students in the past 10 years have published an NBER working paper and are now at a Norwegian institution?

@@

Many well-known survey organizations impose redistribution and usage restrictions that are not consistent with the OKD definition. Many such redistribution restrictions apply to datasets with more detailed personal information collected through surveys, and are meant to ensure continued compliance with ethical rules of behavior, often with informed consent agreements by survey participants and local privacy laws. Notable examples include  PSID \parencite{institute_for_social_research_panel_2024}, World Value Survey \parencite{haerpfer_world_2024}, \ac{DHS} Program \parencite{dhs_program_demographic_2024}, German Socio-Economic Panel \parencite{goebel_german_2019,goebel_socio-economic_2024}, all of which have broad (cost-free) usage, conditional on registration and compliance with usage restrictions.\footnote{I come back to the case of the \ac{DHS} Program in Conclusion.} Table~\ref{tab:restrictions} shows a few examples.

@@

The last three rows of  Table~\ref{tab:restrictions}, however, go further. Researchers using SOEP data must satisfy certain geographical requirements, such as presence of the researcher in the countries covered by \ac{GDPR}, otherwise they can only access a modified version of the data. Similar restrictions apply to the restricted-use data of many other surveys, for instance, the US-based \ac{NLSY}. Users of restricted PSID data must comply with even more stringent rules: secure approval from their institution's ethics board, and use of secure computing environments. Clearly, these restrictions will inhibit broader use of the data per se. Yet these restrictions are not globally very restrictive: There are in Economics alone \economicsgrads{} new US-based researchers being given the ability to request access to geo-restricted NLSY data every year: newly minted Ph.Ds, as noted earlier. A similar number in the EU likely obtain the right to access the geo-restricted SOEP data as well. Despite the restrictions on the use of SOEP data, there are currently \soeppapers{} papers that in some fashion have used the data.\footnote{Source: \url{https://www.diw.de/en/diw_01.c.789503.en/publications_based_on_soep_data__soeplit.html} accessed on 2025-02-08.} 
The PSID bibliography\footnote{Source: \url{https://psidonline.isr.umich.edu/publications/Bibliography/search.aspx} accessed on 2025-02-08} lists over 1,300 dissertations and over 5,300 articles.
In the space depicted in Figure~\ref{fig:nxt}, access is very much towards the left of the figure.%
%
\footnote{In fact, authors sometimes forget to abide by the rules for these datasets. As AEA Data Editor, I get notified via ``take-down requests'' from data providers 12-15 times per year, including in 2024 from the PSID, and have posted information on how to achieve compliance, at least in some cases, at \url{https://aeadataeditor.github.io/posts/2024-11-01-psid-requests}. Most of the cases affect papers published prior to my tenure, because I do alert authors to  data use agreement violations that I am aware of. Ultimately, however, it is the authors' obligation to remain compliant with such data use agreements.}.

\paragraph{Organic data:} However, most economists, when asked about data subject to restrictions, will think of ``proprietary data,'' a term often applied to any data that may be subject to restrictions of use and re-use. Access to most administrative data is typically not ``open'' in the sense of OKD, but are they open enough, given the privacy concerns that are attached to these data? Many have argued that access is not broad enough \citep{card_expanding_2010,einav_economics_2014}, while acknowledging the difficulty of addressing privacy and security of the data at scale. \citet{abowd_economic_2018} discuss the challenge of making the choice of between accuracy of (public) statistics and data, and the privacy loss inherent in doing so. \citet{nagaraj_improving_2020} argue that more openness improves scientific progress, and \citet{nagaraj_how_2023} study this in the context of the US system for providing access (\ac{FSRDC}). They note that 4\% of US-based empirical authors have had some access to the FSRDC system. I am not aware of similar studies for other countries, such as France \citep[the equivalent system is the \ac{CASD}, ][]{gadouche_centre_2019} or Canada \citep{currie2015social}. In absolute terms, these networks host several hundred researchers every year. As per (CENSUS), 2,084 researchers have had access to projects involving Census Bureau data since (YEAR).\footnote{\url{https://labordynamicsinstitute.github.io/fsrdc-external-census-projects/}} \citet{nagaraj_how_2023} mention 861 papers. Similar numbers can be obtained for the French (xxx researchers and 417 publications)
% https://www.casd.eu/toutes-les-publications/liste/0/20/ 
% 20 pages à 20 publications et 1 à 17 = 417
and Canadian (2201  active researchers as of January 2025, and 3245 papers published between 2000 and May 2024)\footnote{Provided by Grant Gibson, CRDCN, on February 10, 2025.} networks.\footnote{The \citet{nagaraj_how_2023} number only includes papers published by economists. All other numbers are counts of researchers and publications in all disciplines, and include non-economists.} The number of publications from access to these networks is smaller in absolute terms then those from PSID and SOEP, though likely higher in impact \citep{nagaraj_how_2023}.

\paragraph{Primary data collection:} The discussion of choices made by survey organizations should in principle be applicable when smaller teams of economists, not entire survey organizations, do the primary data collection. Similar to survey organizations, such teams have to balance the privacy of their respondents with the benefits of open science, in particular the broader knowledge to be gained from open access to the data. Many, so it would seem, provide much of the data in replication packages, subject  to de-identification \citep[see ][ for examples]{kopper_j-pal_2020,bjarkefur_development_2021}, though typically not with stronger disclosure avoidance measures similar to those employed by statistical agencies and larger survey institutions \citep[for a brief discussion of the issues and one possible solution, see ][]{mukherjee_assessing_2023}. For research teams, ethics boards and \acp{IRB} have a role to play \citep{grant_opinion_2019}, with some arguing very strongly that greater availability to others (though not blind publication of all data) is required in order to maximize the societal benefits that are the \textit{quid pro quo} for the respondents' consent to their privacy being invaded \citep{meyer_practical_2018,grant_opinion_2019}. Making such data as broadly available, while respecting the privacy of respondents, is precisely what open access to such data promises, \textit{modulo} appropriate access restrictions or data use agreements similar to those outlined in Table~\ref{tab:restrictions}. In general, however, primary data collections do not have access to robust third-party systems that would allow for access similar to the access required by PSID and similar organizations, situated between no access and fully public access. Thus, while access may be requested in ad-hoc fashion via the original authors, this is known to be fraught with problems \citep{watson_many_2022,gabelica_many_2022}. An ideal scenario would see researchers deposit the data they collected in third-party repositories, which then handle issues such as verifying ethics approval and secure access mechanisms. Some full-service repositories, such as \ac{ICPSR} or various national archives, offer such deposits, though they are rarely used by individual researchers in economics.
%
% Swedish example
% Das paper ist
One example are the data collected and used by \citet{ahrsjo_identity_2024}. The data were collected from public information from the Stockholm District Court. However, in combining individuals' information into a database, the result was subject to \ac{GDPR}, and could not be included in the replication package \citep{ahrsjo_code_2024}. 
%Ahrsjö, Ulrika, Susan Niknami, and Mårten Palme. “Identity in Court Decision-Making.” American Economic Journal: Economic Policy 16, no. 4 (November 2024): 142–64. https://doi.org/10.1257/pol.20220802.
%
% Ahrsjö, Ulrika, Niknami, Susan, and Palme, Mårten. Code for Identity in Court Decision-Making. Nashville, TN: American Economic Association [publisher], 2024. Ann Arbor, MI: Inter-university Consortium for Political and Social Research [distributor], 2024-10-07. https://doi.org/10.3886/E193364V1
% Im Kontext, die Autoren bauen einen "data set created from court documents from the Stockholm District Court." Den dürfen sie aber nicht, laut ihrer Uni, wegen GDPR veröffentlichen. Also haben sie ihn beim "Swedish National Data Service" hinterlegt:
%
% Ahrsjö, U., Niknami, S., & Palme, M. (2023). Data för: Identity in Court Decision-Making (Version 1) [Dataset]. Stockholms universitet. Tillgänglig via: https://doi.org/10.58141/27j7-8606
However, the authors were able to deposit their pseudoymous data at the Svensk nationell datatjänst (Swedish National Data Service), as a restricted dataset \citep{ahrsjo_data_2023}. The SNDS will verify compliance with Swedish law (e.g., \ac{GDPR}), without requiring future involvement of the original collectors of the data.

How onerous are access restrictions in general? In my work as Data Editor, it matters less than it would seem at first blush. The AEA asks that authors, when submitting, provide some information on how restrictive the data used in the article are, and whether the authors are able to provide the data editor with a not-for-publication copy, for the purpose of verification.\footnote{\href{https://www.aeaweb.org/journals/forms/data-code-availability}{Data and Code Availability Form}} Table~\ref{tab:categories} lists the four levels of restrictions, with some guidance on how to classify specific data. Access via research data centers, described earlier, generally fit within the `moderately difficult to obtain` category. Many private-sector datasets, relying on personal interactions with individuals within the company, fall into the `very difficult to obtain', because others may never be able to obtain them. Also in these lists are access-restricted data that have been discontinued, and are thus no longer available to anybody, such as Zillow's  ZTRAX program, terminated in 2021 \citep{zillow_zillows_2021}. Authors can list multiple categories, and we review these, possibly adjusting them before we record them in our internal database.

@@

Figure~\ref{fig:access} lists the distribution of the four levels of restrictions plus  the no-restriction category, for the \aerpapers{} AER articles in \aeryear{}, interacted with whether the authors offered to share the files with the data editor.\footnote{Not all offers to share data are taken up, depending on available resources and time. Appendix Table~\ref{tab:access_provision} shows the same numbers.}  
%
Past statistics note that about 40\% of articles rely on data that is not freely available \parencite{herbert_reproduce_2024,einav_economics_2014}. 
%
In \aeryear{}, slightly more than 50\% of all articles had no data restrictions at all (included the data in the replicatin package), with another 19\% being `very easy to obtain'. The combined percentage of 70\% is higher than the earlier long-run average.  More importantly, out of the 39 data sources that are difficult to obtain, about a third could actually be shared with the data editor. It is also worthwhile highlighting that the `moderately difficult to obtain' tends to be data from research data center networks that have fairly robust mechanisms of access, so that in principle and in practice, many researchers do have access to the same data used in these articles, even when the data editor may not have the time to access these data, and when it is not a legal option for the authors to share the data with the data editor. Arguably, the only data sources that are of immediate concern are data where access is uniquely attributable to the authors of the paper, or where access is no longer possible.

\input{figure3}

@@

Concern should be raised, however, for any data that is subject to a gatekeeping mechanism, however objective, impartial, and lightweight it may be at this time, as such mechanisms can and do disappear.  A particular striking example, as of the writing of this article, is access to data from the USAID-funded \ac{DHS} program. Data access of such data was classified as `moderately easy to obtain' in Table~\ref{tab:categories} until February 2025, as it took only a day or two to obtain access subject to a lightweight data use agreement, a process my team at the AEA regularly went through to obtain data used by other researchers. In February 2025, the second Trump administration shut down USAID and the DHS program, with very little notice. While the system remained (for a time) accessible to those with prior access permission, no further access requests were accepted. That turns it into a `very difficult to obtain' dataset.\footnote{As of March 2025, efforts are underway to preserve the DHS program data, for instance via IPUMS, and to resuscitate both the access to historical data, as well continue funding new surveys.} These events are a note of caution that any kind of redistribution restriction may very well negatively impact future availability to the research community.

\printbibliography[title={References}]

\appendix
\appendixpage
